---
title: "Personality Analyses Final"
output: html_notebook
---

# Preparing libraries and clearing world environment
```{r}
rm(list = ls())

library(tidyverse)
```

# Loading ethogram
```{r}
# loading required files
personality_focal <- read.csv("~/Savolainen Lab/SSB/Data/Personality/Updated_Chloe/clean_focal_follow_data.csv")

ethogram <- tribble(
  ~action,                  ~Category,

  # Assertive
  "(give) Aggression",        "Assertive",
  "Display",                  "Assertive",
  "Open-mouth threat",        "Assertive",
  "Supplant",                 "Assertive",

  # Submissive
  "(receive) Aggression",     "Submissive",
  "Cower",                    "Submissive",
  "Lip smack",                "Submissive",
  "Make room",                "Submissive",
  "Silent bared-teeth face",  "Submissive",

  # Affiliative
  "Brow flash",               "Affiliative",
  "Cooperative infant care",  "Affiliative",
  "(give) Food share",        "Affiliative",
  "(receive) Food share",     "Affiliative",
  "(give) Groom",             "Affiliative",
  "(receive) Groom",          "Affiliative",
  "Hold others",              "Affiliative",
  "Initiate contact",         "Affiliative",
  "Muzzle contact",           "Affiliative",
  "Play",                     "Affiliative",
  "Teeth chatter",            "Affiliative",

  # Sexual
  "(give) Genital stimulation","Sexual",
  "(receive) Genital stimulation","Sexual",
  "(give) Mount",             "Sexual",
  "(receive) Mount",          "Sexual",

  # Exploratory
  "Forage",                   "Exploratory",
  "Object interaction",       "Exploratory",
  "Other species interaction","Exploratory",

  # Neutral
  "Feed",                     "Neutral",
  "Rest",                     "Neutral",

  # Locomotion
  "Locomotion",               "Locomotion",

  # Approach/Avoidance
  "Approach",                 "Approach",
  "Avoid",                    "Avoidance",
  "Reject",                   "Avoidance",

  # Passive contact
  "Passive body contact",     "Passive contact",
  "Stationary in proximity",  "Passive contact",

  # Alone
  "Stationary alone",         "Alone duration",

  # Vigilance
  "Stare",                    "Vigilance",
  "Vigilant scan",            "Vigilance",

  # Self-directed behaviour
  "Masturbate",               "Self-directed behaviour",
  "Self-groom",               "Self-directed behaviour",

  # Vocalisation
  "Vocalisation",             "Vocalisation",

  # Out of sight
  "Out of sight",             "Out of sight"
)

# Check for any behaviours in the data not covered by the ethogram
unmapped <- personality_focal %>%
  anti_join(ethogram, by = "action") %>%
  distinct(action)

if (nrow(unmapped) > 0) {
  message("Unmapped behaviours: ", paste(unmapped$action, collapse = ", "))
} else {
  message("All behaviours successfully mapped.") }

```

# Analysing behavioural repeatability

## Checking number of years each monkey appears in
All monkeys only appear once across dataset 
```{r}
# checking the number of years each monkey appears in
personality_focal_mon <- personality_focal %>% 
  group_by(actor_id) %>% 
  summarise(
    n_years = n_distinct(year),
    years = paste(sort(unique(year)), collapse = ",")
  ) %>% 
  arrange(desc(n_years))
personality_focal_mon
```

## Filtering behaviour
We need to filter rare behaviour as they are not biologically significant and meaningful for our analysis; and they will skew the PCA. This is due to low variance, and introduction of instability into analyses (too many zeros)

### Removing behaviours not pre-defined in ethogram
These behaviours are not biologically significant
```{r}
removed_behaviours <- personality_focal %>%
  anti_join(ethogram, by = "action") %>%
  distinct(action)

removed_behaviours

personality_focal_filtered <- personality_focal %>%
  semi_join(ethogram, by = "action")

```

### Recreating proportions
```{r}
personality_prop_by_focal <- personality_focal_filtered %>%
  group_by(actor_id, focal_number, action) %>%
  summarise(
    action_duration = sum(action_duration, na.rm = TRUE),
    total_duration = first(total_duration),
    .groups = "drop"
  ) %>%
  mutate(
    action_proportion = action_duration / total_duration
  )
```


### Checking rare behaviour
First, checking number of counts per behaviour
```{r}
behaviour_counts <- personality_prop_by_focal %>%
  group_by(action) %>%
  summarise(
    n_obs = n(),
    n_monkeys = n_distinct(actor_id),
    .groups = "drop"
  ) %>%
  arrange(n_obs)

behaviour_counts
```

### Removing rare behaviours
```{r}
keep_behaviours <- behaviour_counts %>%
  filter(
    n_obs >= 5,
    action != "Out of sight" # also removing out of sight
  ) %>%
  pull(action)

personality_repeatability <-
  personality_prop_by_focal %>%
  filter(action %in% keep_behaviours)

removed_stat <- behaviour_counts %>%
  filter(!(action %in% keep_behaviours)) %>%
  arrange(n_obs)
removed_stat

```
### Creating new object with proportions
# Merging and calculating average proportion of each action per monkey
This personality matrix excludes all rare behaviours, "out of sight" and behaviours not found in the ethogram
```{r}
# calculating proportions per focal session
personality_prop_by_focal <- personality_repeatability %>%
  group_by(actor_id, focal_number, action) %>%
  summarise(
    action_duration = sum(action_duration, na.rm = TRUE),
    total_duration = first(total_duration), .groups = "drop"
  ) %>%
  mutate(
    action_proportion = action_duration / total_duration
  )

# checking the total proportions per focal session
focal_proportion_summary <- personality_prop_by_focal %>%
  group_by(actor_id, focal_number) %>%
  summarise(
    total_proportion = sum(action_proportion, na.rm = TRUE),
    .groups = "drop"
  )

# number of focal sessions per monkey
n_focals_per_monkey <- personality_prop_by_focal %>%
  distinct(actor_id, focal_number) %>%
  count(actor_id, name = "n_focals")

# mean proportion per monkey per action
personality_summary <- personality_prop_by_focal %>%
  group_by(actor_id, action) %>%
  summarise(
    total_proportion = sum(action_proportion, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  left_join(n_focals_per_monkey, by = "actor_id") %>%
  mutate(
    mean_proportion = total_proportion / n_focals
  )

# checking total proportions per monkey
monkey_proportion_summary <- personality_summary %>%
  group_by(actor_id) %>%
  summarise(
    total_proportion = sum(mean_proportion, na.rm = TRUE),
    .groups = "drop"
  )

# saving file
write.csv(personality_summary, "~/Savolainen Lab/SSB/Data/Personality/new_filtered_personality_summarised.csv")
```

### Checking variance of behaviours
```{r}
personality_matrix <- personality_summary %>%
  dplyr::select(actor_id, action, mean_proportion) %>%
  pivot_wider(
    names_from = action,
    values_from = mean_proportion,
    values_fill = 0
  ) %>%
  rename(Animal_ID = actor_id)


# checking variance of behaviours
behaviour_variance <- personality_matrix %>%
  summarise(across(everything(), ~ var(.x, na.rm = TRUE))) %>%
  pivot_longer(
    everything(),
    names_to = "Behaviour",
    values_to = "Variance"
  ) %>%
  arrange(Variance)
behaviour_variance
```

## Running repeatability function
### Creating a new dataframe filling unobserved behaviours as 0.
```{r}
all_focals <- personality_focal_filtered %>%
  distinct(actor_id, focal_number)

all_combinations <- tidyr::crossing(
  all_focals,
  action = keep_behaviours
)

personality_repeatability <- all_combinations %>%
  left_join(
    personality_prop_by_focal,
    by = c("actor_id", "focal_number", "action")
  ) %>%
  mutate(
    action_proportion = replace_na(action_proportion, 0)
  )
```

### Checking assumptions of normality for each behaviour
#### Checking histograms for each behaviour
```{r}
library(ggplot2)

ggplot(
  personality_repeatability,
  aes(x = action_proportion)
) +
  geom_histogram(bins = 30) +
  facet_wrap(~ action, scales = "free") +
  theme_bw()
```


#### Checking proportions of zeros for behaviour
```{r}
zero_summary <- personality_repeatability %>%
  group_by(action) %>%
  summarise(
    n = n(),
    n_zero = sum(action_proportion == 0),
    prop_zero = mean(action_proportion == 0),
    .groups = "drop"
  ) %>%
  arrange(desc(prop_zero))

zero_summary
```


### Checking sampling effort for repeatability checks
```{r}
repeatability_check <- personality_repeatability %>%
  group_by(action, actor_id) %>%
  summarise(
    n_focals = n(),
    .groups = "drop"
  ) %>%
  group_by(action) %>%
  summarise(
    n_monkeys = n(),
    min_focals = min(n_focals),
    median_focals = median(n_focals),
    max_focals = max(n_focals),
    .groups = "drop"
  )

repeatability_check
```


### Running repeatability function
```{r, results = 'hide', message=FALSE, warning=FALSE}
library(rptR)
library(dplyr)
library(purrr)

behaviour_repeatability <- personality_repeatability %>%
  group_split(action) %>%
  map_df(function(dat){

    fit <- tryCatch(
      rpt(
        action_proportion ~ (1 | actor_id),
        grname = "actor_id",
        data = dat,
        datatype = "Gaussian",
        nboot = 1000,
        npermut = 1000
      ),
      error = function(e) return(NULL)
    )

    if (is.null(fit)) {
      return(tibble(
        Behaviour = unique(dat$action),
        Repeatability = NA,
        LowerCI = NA,
        UpperCI = NA,
        P = NA
      ))
    }

   tibble(
  Behaviour = unique(dat$action),
  Repeatability = fit$R$actor_id,
  LowerCI = fit$CI_emp$`2.5%`,
  UpperCI = fit$CI_emp$`97.5%`,
  P = fit$P$P_permut,
  LRT_P = fit$P$LRT_P,
  N_obs = fit$nobs,
  N_monkeys = unname(fit$ngroups)
)
  })

View(behaviour_repeatability)
write.csv(behaviour_repeatability, "~/Savolainen Lab/SSB/Data/Personality/repeatability_analysis.csv")
```

# Loading the required rank and rate files
```{r}
rank_file <- read.csv("~/Savolainen Lab/SSB/Data/Rank/rank_all_personality.csv")

rate_file <- read.csv("~/Savolainen Lab/SSB/Data/updatedssbrates.csv") %>% 
  mutate(
    Animal_ID = dplyr::recode(Animal_ID,
                            "5E1" = "5EE1",
                            "6E6" = "6EE6"))

age <- read.csv("~/Savolainen Lab/Age/age_behaviour.csv") %>% 
  dplyr::select(-"X")

ag <- read.csv("~/Savolainen Lab/SSB/Data/Aggression/Ag_lifetime.csv") %>% 
    mutate(
    Animal_ID = dplyr::recode(Animal_ID,
                            "5E1" = "5EE1",
                            "6E6" = "6EE6"))

rank_rate <- left_join(rank_file, rate_file, by = "Animal_ID")

rank_rate <- left_join(rank_rate, age,
                       by = "Animal_ID")

rank_rate <- left_join(rank_rate, ag, 
                       by = "Animal_ID") %>% 
  dplyr::select(c("Animal_ID", "age_years", "DS", "DS_norm", "adjusted_SSB", "adjusted_SSB_giver", "adjusted_SSB_receiver", "adjusted_ag_giver", "adjusted_ag_receiver", "adjusted_ag_total"))

# replacing NAs in aggression with 0, as present SSB rates means they have been observed in scans but not observed with aggression
rank_rate <- replace_na(
  rank_rate,
  list(
    adjusted_ag_giver = 0,
    adjusted_ag_receiver = 0,
    adjusted_ag_total = 0
  )
)
write.csv(rank_rate, "~/Savolainen Lab/SSB/Data/rank_rates_updated.csv")
```

# PCA (Varimax rotation)
## Preparing for PCA

```{r}
library(psych)
library(tibble)

behaviour_data <- personality_matrix %>%
  column_to_rownames("Animal_ID")

behaviour_z <- scale(behaviour_data)

cor_matrix <- cor(behaviour_z)
cortest.bartlett(cor_matrix,
                 n = nrow(behaviour_z))
```

## Running parallel analysis
This determines how many components are relevant for analysis. 3 components suggested here.
```{r}
parallel <- fa.parallel(
  behaviour_z,
  fa = "pc",
  n.iter = 1000
)
```

## Kaiser correction
This also examines how many components to analyse. 14 components suggested here
```{r}
eigenvalues <- eigen(cor_matrix)$values
eigenvalues
sum(eigenvalues > 1)
```

## Carrying out PCA
Loading of 0.30 for full loading
```{r}
pca <- principal(
  behaviour_z,
  nfactors = 3,
  rotate = "varimax",
  scores = TRUE
)

print(
  pca,
  cutoff = 0.30,
  sort = TRUE
)

# Variance explained per axis
variance_explained <- pca$Vaccounted
variance_explained

# extracting pca scores
pca_scores <- as.data.frame(pca$scores) %>%
  rownames_to_column("Animal_ID")

# combining pcoa_scores with rank and rates
pca_scores_rankrate <- left_join(pca_scores, rank_rate, by = "Animal_ID")

write.csv(pca_scores_rankrate, "~/Savolainen Lab/SSB/Data/Personality/pca_scores_rankrate.csv")
```

### PCA (cut-off = 0.40)
```{r}
pca <- principal(
  behaviour_z,
  nfactors = 3,
  rotate = "varimax",
  scores = TRUE
)

print(
  pca,
  cutoff = 0.40,
  sort = TRUE
)
```

## Extracting loadings for each RC
```{r}
loadings_df <- as.data.frame(unclass(pca$loadings)) %>%
  rownames_to_column("Behaviour")
loadings_df

write.csv(loadings_df, "~/Savolainen Lab/SSB/Data/Personality/pca_loadings_all.csv")

# RC1
RC1_loadings <- loadings_df %>%
  filter(abs(RC1) >= 0.40) %>%
  arrange(desc(abs(RC1))) %>% 
  dplyr::select(c("Behaviour", "RC1"))

# RC2
RC2_loadings <- loadings_df %>%
  filter(abs(RC2) >= 0.40) %>%
  arrange(desc(abs(RC2))) %>% 
  dplyr::select(c("Behaviour", "RC2"))

# RC3
RC3_loadings <- loadings_df %>%
  filter(abs(RC3) >= 0.40) %>%
  arrange(desc(abs(RC3))) %>% 
  dplyr::select(c("Behaviour", "RC3"))

RC1_loadings
RC2_loadings
RC3_loadings

write.csv(RC1_loadings, "~/Savolainen Lab/SSB/Data/Personality/rc1_loading.csv")
write.csv(RC2_loadings, "~/Savolainen Lab/SSB/Data/Personality/rc2_loading.csv")
write.csv(RC3_loadings, "~/Savolainen Lab/SSB/Data/Personality/rc3_loading.csv")
```