---
title: "For Thesis: Personality-Rate Linear Models"
output: html_notebook
---

Here, we carry out linear models for our PCA scores.

# Preparing world environment
```{r}
rm(list = ls())

library(tidyverse)
library(ggeffects)
library(ggplot2)
library(patchwork)
library(grid)

pca_scores <- read.csv("~/Savolainen Lab/SSB/Data/Personality/pca_scores_rankrate.csv")

# removing NA values
pca_scores <- pca_scores %>% 
  drop_na(
  DS)
```

## Adding group and year, so they can be used
```{r}
year <- read.csv("~/Savolainen Lab/SSB/Data/Personality/Updated_Chloe/focals_used_forrank.csv") %>% 
  rename(Animal_ID = ID)

group <- read.csv("~/Savolainen Lab/SSB/Data/Rank/rank_all_personality.csv") %>% 
  dplyr::select(c("Animal_ID", "Group"))

pca_scores <- left_join(pca_scores, year, by = "Animal_ID")
pca_scores <- left_join(pca_scores, group, by = "Animal_ID")
```

## Removing duplicate rows
```{r}
pca_scores <- pca_scores %>%
  dplyr::select(-X) %>%     # Remove the unnecessary row-number column
  distinct(Animal_ID, .keep_all = TRUE)
```

# Checking descriptive statistics
```{r}
mean(pca_scores$age_years)
sd(pca_scores$age_years)
range(pca_scores$age_years)


mean(pca_scores$DS)
sd(pca_scores$DS)
range(pca_scores$DS)

table(pca_scores$Group)
```

# Linear Models:
## Making sure all values are numeric and factors as factors
```{r}
pca_scores <- pca_scores %>%
  mutate(across(
    c(RC1,
      RC2,
      RC3,
      DS,
      DS_norm,
      adjusted_SSB,
      adjusted_SSB_giver,
      adjusted_SSB_receiver,
      age_years),
    as.numeric
  ))

pca_scores <- pca_scores %>% 
  mutate(across(
    c(Year,
      Group),
    as.factor)
  )
```

## Dominance Rank ~ RC1 (Social Confidence)
```{r}
# rank only
lm_rc1 <- lm(RC1 ~ DS + age_years + Group + Year, data = pca_scores)
summary(lm_rc1)

# rank and group interaction
lm_rc1_int <- lm(RC1 ~ DS * Group + age_years + Year, data = pca_scores)
summary(lm_rc1_int)
```

## Rank ~ RC2 (Assertiveness)
```{r}
# rank only
lm_rc2 <- lm(RC2 ~ DS + age_years + Group + Year, data = pca_scores)
summary(lm_rc2)

# rank group int
lm_rc2_int <- lm(RC2 ~ DS * Group + age_years + Year, data = pca_scores)
summary(lm_rc2_int)
```

## Rank ~ RC3 (Evasiveness)
```{r}
# rank only
lm_rc3 <- lm(RC3 ~ DS + age_years + Group + Year, data = pca_scores)
summary(lm_rc3)

# rank, interaction with group
lm_rc3_int <- lm(RC3 ~ DS * Group + age_years + Year, data = pca_scores)
summary(lm_rc3_int)
```

# Checking model assumptions
## RC1
Diagnostic plots look okay
### RC1 (Rank only)
```{r}
# Fit model
lm_test1 <- lm(RC1 ~ DS + age_years + Group + Year, data = pca_scores)

summary(lm_test1)

# Diagnostic plots
par(mfrow = c(2, 2))
plot(lm_test1)
par(mfrow = c(1, 1))
```

### RC1 ~ Rank, Group Interaction
```{r}
# Fit model
lm_test1a <- lm(RC1 ~ DS * Group + age_years + Year, data = pca_scores)

summary(lm_test1a)

# Diagnostic plots
par(mfrow = c(2, 2))
plot(lm_test1a)
par(mfrow = c(1, 1))
```

## RC2
### RC2 (rank only)
```{r}
# Fit model
lm_test2 <- lm(RC2 ~ DS + age_years + Group + Year, data = pca_scores)

summary(lm_test2)

# Diagnostic plots
par(mfrow = c(2, 2))
plot(lm_test2)
par(mfrow = c(1, 1))
```

### RC2 (rank, group interaction)
```{r}
# Fit model
lm_test2c <- lm(RC2 ~ DS * Group + age_years + Year, data = pca_scores)

summary(lm_test2c)

# Diagnostic plots
par(mfrow = c(2, 2))
plot(lm_test2c)
par(mfrow = c(1, 1))
```

## RC3
### RC3 (rank only)
```{r}
# Fit model
lm_test3 <- lm(RC3 ~ DS + age_years + Group + Year, data = pca_scores)

summary(lm_test3)

# Diagnostic plots
par(mfrow = c(2, 2))
plot(lm_test3)
par(mfrow = c(1, 1))
```


To justify the use of HC3, cite:
1.Long, J.S. & Ervin, L.H. (2000).

Using Heteroscedasticity Consistent Standard Errors in the Linear Regression Model.

2. Hayes, A.F. & Cai, L. (2007).

Using heteroskedasticity-consistent standard error estimators in OLS regression: An introduction and software implementation.


The American Statistician, 54, 217–224.

### RC3 (rank,Group interaction)
```{r}
# Fit model
lm_test3c <- lm(RC3 ~ DS * Group + age_years + Year, data = pca_scores)

summary(lm_test3c)

# Diagnostic plots
par(mfrow = c(2, 2))
plot(lm_test3c)
par(mfrow = c(1, 1))
```

# Plotting linear models of dominance rank on personality effects
## Finding y-axis scale
```{r}
pred_rc1 <- predict_response(
  lm_rc1,
  terms = "DS",
  margin = "empirical"
)

pred_rc2_int <- predict_response(
  lm_rc2_int,
  terms = c("DS", "Group"),
  margin = "empirical"
)

pred_rc3 <- predict_response(
  lm_rc3, 
  terms = "DS",
  margin = "empirical"
)

y_lim <- range(
  pca_scores$RC1,
  pca_scores$RC2,
  pca_scores$RC3,
  pred_rc1$conf.low,
  pred_rc1$conf.high,
  pred_rc2_int$conf.low,
  pred_rc2_int$conf.high,
  pred_rc3$conf.low,
  pred_rc3$conf.high,
  na.rm = TRUE
)

x_lim <- range(pca_scores$DS, na.rm = TRUE)

theme_personality <-
  theme_classic(base_size = 20) +
  theme(
    axis.title = element_text(
      size = 20,
      face = "plain"
    ),
    axis.text = element_text(
      size = 17
    ),
    axis.ticks.length = unit(0.22, "cm"),
    legend.title = element_text(
      size = 18,
      face = "bold"
    ),
    legend.text = element_text(
      size = 17
    ),
    plot.tag = element_text(
      size = 20,
      face = "bold"
    )
  )
```


## RC1
```{r}
p_rc1 <-
  ggplot(
    pca_scores,
    aes(DS, RC1)
  ) +

  geom_point(
    alpha = 0.65,
    size = 3
  ) +

  geom_ribbon(
    data = pred_rc1,
    aes(
      x = x,
      ymin = conf.low,
      ymax = conf.high
    ),
    inherit.aes = FALSE,
    alpha = 0.08,
    colour = NA
  ) +

  geom_line(
    data = pred_rc1,
    aes(
      x = x,
      y = predicted
    ),
    inherit.aes = FALSE,
    linewidth = 2
  ) +

  scale_x_continuous(
    limits = x_lim
  ) +

  scale_y_continuous(
    limits = y_lim
  ) +

  labs(
    x = NULL,
    y = "Social Confidence"
  ) +

  theme_personality

p_rc1
```

## RC2
```{r}
p_rc2_int <- ggplot(pca_scores,
                    aes(DS, RC2)) +

  # Observed data
  geom_point(
    aes(colour = Group),
    alpha = 0.65,
    size = 3
  ) +

  # 95% confidence intervals
  geom_ribbon(
    data = pred_rc2_int,
    aes(
      x = x,
      ymin = conf.low,
      ymax = conf.high,
      fill = group,
      group = group
    ),
    inherit.aes = FALSE,
    alpha = 0.08,
    colour = NA,
    show.legend = FALSE
  ) +

  # Predicted lines
  geom_line(
    data = pred_rc2_int,
    aes(
      x = x,
      y = predicted,
      colour = group
    ),
    inherit.aes = FALSE,
    linewidth = 2.0
  ) +

    scale_x_continuous(
    limits = x_lim
  ) +
  
  scale_y_continuous(
    limits = y_lim
  ) +
  
  scale_colour_manual(values = c(
    R = "#D55E00",
    V = "#0072B2"
  )) +

  scale_fill_manual(values = c(
    R = "#D55E00",
    V = "#0072B2"
  )) +

  labs(
    x = NULL,
    y = "Assertiveness",
    colour = "Group"
  ) +

theme_personality

p_rc2_int

# Saving pplot
ggsave(
  filename = "~/Savolainen Lab/Figures/RC2_int_new.tiff",
  plot = p_rc2_int,
  width = 6,
  height = 5,
  dpi = 600,
  compression = "lzw"
)

p_rc2_int <- p_rc2_int +
  theme(
    legend.position = "none"
  )
```

## RC3
```{r}
p_rc3 <- ggplot(pca_scores, aes(DS, RC3)) +

  geom_point(alpha = 0.65, size = 3) +

  geom_ribbon(
    data = pred_rc3,
    aes(x = x, ymin = conf.low, ymax = conf.high),
    inherit.aes = FALSE,
    alpha = 0.08,
    colour = NA
  ) +

  geom_line(
    data = pred_rc3,
    aes(x = x, y = predicted),
    inherit.aes = FALSE,
    linewidth = 2.0
  ) +
  
 scale_x_continuous(
    limits = x_lim
  ) +

  scale_y_continuous(
    limits = y_lim
  ) +
  
  labs(
    x = "David's Score",
    y = "Evasiveness"
  ) +

theme_personality

p_rc3

ggsave(
  filename = "~/Savolainen Lab/Figures/RC3_new.tiff",
  plot = p_rc3,
  width = 6,
  height = 5,
  dpi = 600,
  compression = "lzw"
)
```

## Combining all three
```{r}
combined_plot <-
  (
    (p_rc1 +
       theme(
         legend.position = "none"
       )) /

    (p_rc2_int +
       theme(
         legend.position = "top",
         legend.justification = "center"
       ) +
       guides(
         colour = guide_legend(
           override.aes = list(
             shape = 20,
             size = 5,
             linewidth = 2.8
           )
         ),
         fill = "none"
       )) /

    (p_rc3 +
       theme(
         legend.position = "none"
       ))
  ) +
  plot_annotation(
    tag_levels = "A"
  ) &
  theme(
    plot.tag = element_text(
      size = 20,
      face = "bold"
    )
  )

combined_plot

ggsave(
  filename = "~/Savolainen Lab/Figures/RC1_RC2_RC3_combined_new.tiff",
  plot = combined_plot,
  width = 10,
  height = 12,
  dpi = 600
)
```

