---
title: "For Thesis: Targetted Genera Analysis for Genera"
output: html_notebook
---

This notebook uses MaAsLin2 for targetted genera, which is Microbiome Multivariable Associations with Linear Models. The genera has been narrowed down using literature.

# Preparing world environment
```{r}
rm(list = ls())

library(Maaslin2)
library(phyloseq)
library(qiime2R)
library(tidyverse)
library(broom)
```

## Reading files
```{r}
# Change paths if needed
table_qza    <- "~/Savolainen Lab/Microbiome/Qiime/Outputs for PCoA/feature_table_decontam.qza"
taxonomy_qza <- "~/Savolainen Lab/Microbiome/Qiime/Outputs for PCoA/taxonomy_silva.qza"
metadata_tsv <- "~/Savolainen Lab/Microbiome/Qiime/Outputs for PCoA/metadata.tsv"

# Feature table
table <- read_qza(table_qza)$data

# Taxonomy
taxonomy <- read_qza(taxonomy_qza)$data

# Metadata
metadata <- read.table(
  metadata_tsv,
  header       = TRUE,
  sep          = "\t",
  row.names    = 1,
  comment.char = "",
  check.names  = FALSE
) %>%
  tibble::rownames_to_column("SampleID")
```

# Building the Phyloseq Object
```{r}
# OTU table — remove BLANK and CTR samples, taxa are rows in QIIME2 feature tables
exclude_samples <- colnames(table)[grepl("BLANK|CTR", colnames(table), ignore.case = TRUE)]
otu_data <- table[, !colnames(table) %in% exclude_samples]
otu <- otu_table(as.matrix(otu_data), taxa_are_rows = TRUE)

# Split taxonomy into ranks (separator is ";" with no space)
tax_split <- str_split_fixed(taxonomy$Taxon, ";", 7)
colnames(tax_split) <- c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species")
rownames(tax_split) <- taxonomy$Feature.ID

tax  <- tax_table(as.matrix(tax_split))

metadata_ps <- metadata %>%
  tibble::column_to_rownames("SampleID")

meta <- sample_data(metadata_ps)
ps <- phyloseq(otu, tax, meta)

# Filter: keep taxa present in >= 4 samples AND with total abundance >= 20
ps <- filter_taxa(ps, function(x) sum(x >= 1) >= 4 & sum(x) >= 20, prune = TRUE)

ps
```

## Basic checks
```{r}
nsamples(ps)
ntaxa(ps)
sort(sample_sums(ps))
```

## Remove low-depth samples
```{r}
ps_pruned <- prune_samples(sample_sums(ps) > 10000, ps)
ps_pruned
```

# Analyses
## Relative Abundance Transformation
```{r}
ps.rel <- transform_sample_counts(ps_pruned, function(x) x / sum(x))
```

## Preparing metadata file
```{r}
# Behavioural rates
rates <- read.csv("~/Savolainen Lab/SSB/Data/updatedssbrates.csv") %>% 
  mutate(
    Animal_ID = dplyr::recode(Animal_ID,
                            "5E+01" = "5E1")) %>% 
  rename(monkey_id = Animal_ID)
  

# Dominance rank
rank <- read.csv(
  "~/Savolainen Lab/SSB/Data/Rank/rank_all_gut.csv"
) %>%
  rename(monkey_id = Animal_ID) %>%
  mutate(
    monkey_id = dplyr::recode(
      monkey_id,
      "5.00E+01" = "5E1"
    )
  ) %>%
  dplyr::select(-X, -Year)

age <- read.csv("~/Savolainen Lab/Age/age_gut.csv") %>% 
  dplyr::select(-"X") %>% 
  rename(monkey_id = Animal_ID) %>% 
  mutate(
    monkey_id = dplyr::recode(
      monkey_id,
      "5EE1" = "5E1"
    )
  )

# Join behavioural metadata
metadata <- metadata %>%
  left_join(rates, by = "monkey_id") %>%
  left_join(rank, by = "monkey_id")

metadata <- metadata %>% 
  dplyr::select(-c("SSB_Giver", "SSB_Receiver", "DSB_Raw", "SSB_Raw", "X")) %>% 
  left_join(age, by = "monkey_id")

meta_new <- metadata %>% 
  drop_na(DS)
```

```{r}
ps.rel <- prune_samples(meta_new$SampleID, ps.rel)
ps_pruned <- prune_samples(meta_new$SampleID, ps_pruned)

# Keep only metadata for samples retained in the phyloseq object
meta_new <- meta_new %>%
  filter(SampleID %in% sample_names(ps_pruned)) %>%
  arrange(match(SampleID, sample_names(ps_pruned))) %>%
  tibble::column_to_rownames("SampleID")

# Replace sample metadata
sample_data(ps.rel) <- sample_data(meta_new)
sample_data(ps_pruned) <- sample_data(meta_new)

# Check that everything matches
nsamples(ps.rel)
nsamples(ps_pruned)
nrow(meta_new)
```

# Genus-level comparisons
## Agglomerate to genus and melt
```{r}
# Agglomerate counts to genus
ps.genus <- phyloseq::tax_glom(ps_pruned, taxrank = "Genus")

# Extract count matrix
otu <- as(otu_table(ps.genus), "matrix")

# Ensure samples are rows (required by decostand)
if (taxa_are_rows(ps.genus)) {
  otu <- t(otu)
}

# Add pseudocount to avoid zeros
otu <- otu + 1

# CLR transformation
otu.clr <- decostand(otu, method = "clr")

# Put CLR values back into phyloseq
otu_table(ps.genus) <- otu_table(t(otu.clr), taxa_are_rows = TRUE)

# Convert to data frame
genus.df <- psmelt(ps.genus)
```

## Editing metadata
```{r}
meta <- data.frame(sample_data(ps.genus))

meta$Group <- factor(meta$Group)

# Continuous predictors
meta$DS <- as.numeric(meta$DS)
meta$age_years <- as.numeric(meta$age_years)

# SSB variables
meta$adjusted_SSB <- as.numeric(meta$adjusted_SSB)
meta$adjusted_SSB_giver <- as.numeric(meta$adjusted_SSB_giver)
meta$adjusted_SSB_receiver <- as.numeric(meta$adjusted_SSB_receiver)

meta <- meta %>%
  mutate(
    DS_GroupV = ifelse(Group == "V", DS, 0)
  )
```

## Making genus names visible in analyses
```{r}
# Extract genus names
genus_names <- as.character(tax_table(ps.genus)[, "Genus"])

# Replace missing names
genus_names[is.na(genus_names)] <- "Unclassified"

# Make duplicate names unique
genus_names <- make.unique(genus_names)

# Rename columns of otu
colnames(otu.clr) <- genus_names
```

# Immune-associated genera only
## Selecting chosen genera
```{r}
immune_genera <- c(
  # Immune-associated and gut homeostasis (gut health)
  "Agathobacter",
  "Agathobaculum",
  "Alistipes",
  "Anaerostipes",
  "Bacteroides",
  "Blautia",
  "Collinsella",
  "Coprococcus",
  "Dorea",
  "[Eubacterium]",
  "Faecalibacterium",
  "Fusicatenibacter",
  "Intestinimonas",
  "Lachnoclostridium",
  "Lachnospira",
  "Lactobacillus",
  "Ligilactobacillus",
  "Mediterraneibacter",
  "Oscillibacter",
  "Prevotella",
  "Roseburia",
  "Ruminococcus")
```

## Keeping chosen genera
```{r}
chosen_genera_immune <- intersect(immune_genera, colnames(otu.clr))

length(chosen_genera_immune)
chosen_genera_immune

otu.clr.immune <- otu.clr[, chosen_genera_immune, drop = FALSE]
```

##MaAsLin2 Analysis for Rank only
```{r}
fit_rank <- Maaslin2(
    input_data = otu.clr.immune,

    input_metadata = meta,

    output = "MaAsLin2_Genus_ImmuneOnly_Rank",

    fixed_effects = c(
        "DS",
        "Group",
        "age_years"
    ),

    random_effects = NULL,

    normalization = "NONE",

    transform = "NONE",

    analysis_method = "LM",

    standardize = FALSE
)
```

##MaAsLin2 Analysis for Rank, group interaction
```{r}
fit_rank_ssb <- Maaslin2(
    input_data = otu.clr.immune,

    input_metadata = meta,

    output = "MaAsLin2_Genus_ImmuneOnly_Rank_grpint",

    fixed_effects = c(
        "DS",
        "DS_GroupV",
        "Group",
        "age_years"
    ),

    random_effects = NULL,

    normalization = "NONE",

    transform = "NONE",

    analysis_method = "LM",

    standardize = FALSE
)
```


# Pathogenic genera only
## Selecting chosen genera
```{r}
patho_genera <- c(
 # Potentially pathogenic and pro-inflammatory genera  
  "Campylobacter",
  "Clostridium",
  "Desulfovibrio",
  "Enterobacter",
  "Enterococcus",
  "Escherichia-Shigella",
  "Helicobacter",
  "Pseudomonas",
  "Staphylococcus",
  "Streptococcus",
  "Treponema")
```

## Keeping chosen genera
```{r}
chosen_genera_patho <- intersect(patho_genera, colnames(otu.clr))

length(chosen_genera_patho)
chosen_genera_patho

otu.clr.patho <- otu.clr[, chosen_genera_patho, drop = FALSE]
```

##MaAsLin2 Analysis for Rank only
```{r}
fit_rank <- Maaslin2(
    input_data = otu.clr.patho,

    input_metadata = meta,

    output = "MaAsLin2_Genus_PathogenicOnly_Rank",

    fixed_effects = c(
        "DS",
        "Group",
        "age_years"
    ),

    random_effects = NULL,

    normalization = "NONE",

    transform = "NONE",

    analysis_method = "LM",

    standardize = FALSE
)
```

##MaAsLin2 Analysis for Rank, group interaction
```{r}
fit_rank_ssb <- Maaslin2(
    input_data = otu.clr.patho,

    input_metadata = meta,

    output = "MaAsLin2_Genus_PathogenicOnly_Rank_grpint",

    fixed_effects = c(
        "DS",
        "DS_GroupV",
        "Group",
        "age_years"
    ),

    random_effects = NULL,

    normalization = "NONE",

    transform = "NONE",

    analysis_method = "LM",

    standardize = FALSE
)
```



