---
title: "For Thesis: Community Composition Analyses New Hypotheses"
output: html_notebook
---

This notebook will show step by step how the gut microbiome is analysed.
# Preparing the R environment
## Clearing libraries and loading libraries
```{r}
rm(list = ls())

library(qiime2R)
library(phyloseq)
library(tidyverse)
library(vegan)
library(ggrepel)
```

# Data Loading
## Reading Files
```{r}
# Feature table
table <- read_qza("~/Savolainen Lab/Microbiome/Qiime/Outputs for PCoA/feature_table_decontam.qza")

# Taxonomy
taxonomy <- read_qza("~/Savolainen Lab/Microbiome/Qiime/Outputs for PCoA/taxonomy_silva.qza")

# Sequences (optional, needed for tree later)
rep_seqs <- read_qza("~/Savolainen Lab/Microbiome/Qiime/Outputs for PCoA/rep_seqs_decontam.qza")

# Metadata
metadata <- read.table(
  "~/Savolainen Lab/Microbiome/Qiime/Outputs for PCoA/metadata.tsv",
  header = TRUE,
  sep = "\t",
  row.names = 1
)

# Keep original sample IDs
metadata <- metadata %>%
  tibble::rownames_to_column("SampleID")
```

## Preparing metadata
```{r}
# Behavioural rates
rates <- read.csv("~/Savolainen Lab/SSB/Data/updatedssbrates.csv") %>% 
  mutate(
    Animal_ID = dplyr::recode(Animal_ID,
                              "5E+01" = "5E1",
                              "6E+06" = "6EE6")) %>% 
  rename(monkey_id = Animal_ID) %>%
  distinct(monkey_id, .keep_all = TRUE)
  
# Dominance rank
rank <- read.csv(
  "~/Savolainen Lab/SSB/Data/Rank/rank_all_gut.csv"
) %>%
  rename(monkey_id = Animal_ID) %>%
  mutate(
    monkey_id = dplyr::recode(
      monkey_id,
      "5.00E+01" = "5E1"
    )
  ) %>%
  dplyr::select(-X, -Year)

age <- read.csv("~/Savolainen Lab/Age/age_gut.csv") %>% 
  dplyr::select(-"X") %>% 
  rename(monkey_id = Animal_ID) %>% 
  mutate(
    monkey_id = dplyr::recode(
      monkey_id,
      "5EE1" = "5E1"
    )
  )

# Join behavioural metadata
metadata <- metadata %>%
  left_join(rates, by = "monkey_id") %>%
  left_join(rank, by = "monkey_id")

metadata <- metadata %>% 
  dplyr::select(-c("SSB_Giver", "SSB_Receiver", "DSB_Raw", "SSB_Raw", "X")) %>% 
  left_join(age, by = "monkey_id")

metadata <- metadata %>% 
  drop_na(DS)
```

```{r}
# rooted tree loaded separately as it is not ready yet
tree <- read_qza("~/Savolainen Lab/Microbiome/Qiime/Outputs for PCoA/rooted_tree.qza")
```

# Heavy Lifting: Data preparation (OTUs and Taxonomy Analysis)
## Building the phyloseq object
```{r}
# OTU table — remove BLANK and CTR samples, taxa are rows in QIIME2 feature tables
exclude_samples <- colnames(table$data)[grepl("BLANK|CTR", colnames(table$data), ignore.case = TRUE)]

otu_data <- table$data[, !colnames(table$data) %in% exclude_samples]

otu <- otu_table(otu_data, taxa_are_rows = TRUE)

# Taxonomy analysis
tax <- taxonomy$data %>% 
  separate(Taxon,
           into = c("Kingdom","Phylum","Class","Order","Family","Genus","Species"),
           sep = ";", fill = "right") %>% 
  column_to_rownames("Feature.ID") %>% 
  as.matrix()

tax <- tax_table(tax)

metadata <- metadata %>%
  tibble::column_to_rownames("SampleID")

# Metadata
sample_data <- sample_data(metadata)

# Combine
ps <- phyloseq(otu, tax, sample_data)

# Filter: keep taxa present in >= 4 samples AND with total abundance >= 20
ps <- filter_taxa(ps, function(x) sum(x >= 1) >= 4 & sum(x) >= 20, prune = TRUE)

ps

```

## Basic checks on number of samples and taxa
```{r}
nsamples(ps)
ntaxa(ps)
sort(sample_sums(ps))
```

## Remove low-depth samples
```{r}
ps_pruned <- prune_samples(sample_sums(ps) > 10000, ps)
ps_pruned
```

## Inspecting flagged taxa
```{r}
# Load contaminant feature IDs identified by decontam (ran upstream in QIIME2)
contam_ids <- read.table(
  "~/Savolainen Lab/Microbiome/Qiime/output/decontam/full/contaminant_features.tsv",
  header = TRUE, sep = "\t")  %>% 
  pull(feature_id)

# Load the raw (pre-decontam) feature table to assess contaminant read burden
ft_raw <- read.table(
  "~/Savolainen Lab/Microbiome/Qiime/output/decontam/full/feature_table.tsv",
  header = TRUE, sep = "\t", skip = 1, row.names = 1
)

# The 1792 contaminant features are all unclassifiable by SILVA — no taxonomy available
cat("Contaminant features:", length(contam_ids), "\n")
cat("Of these with taxonomy:", sum(contam_ids %in% taxonomy$data$Feature.ID), "\n\n")

# Summarise read counts removed per sample
data.frame(
  sample      = colnames(ft_raw),
  contam_reads = colSums(ft_raw[contam_ids, ]),
  total_reads  = colSums(ft_raw),
  pct_contam   = round(colSums(ft_raw[contam_ids, ]) / colSums(ft_raw) * 100, 2)
)
```

# Beta Diversity Analyses
## Relative Abundance Transformation
```{r}
ps.rel <- transform_sample_counts(ps_pruned, function(x) x / sum(x))
```

## CLR + Aitchison Analyses
```{r}
# Extract OTU table from pruned object, transpose to samples x taxa
otu_clr <- t(as(otu_table(ps_pruned), "matrix"))

# CLR transform (samples are rows)
clr_mat <- decostand(otu_clr, method = "clr", pseudocount = 1)

# Euclidean distance on CLR data (Aitchison distance)
dist_clr <- dist(clr_mat, method = "euclidean")

# PCoA
ord_clr <- cmdscale(dist_clr, eig = TRUE, k = 2)

pcoa_clr_df <- data.frame(
  SampleID = rownames(ord_clr$points),
  Axis.1   = ord_clr$points[, 1],
  Axis.2   = ord_clr$points[, 2]
)

# Join metadata (rownames_to_column needed since metadata uses row names)
pcoa_clr_df <- left_join(
  pcoa_clr_df,
  metadata  %>%  tibble::rownames_to_column("SampleID"),
  by = "SampleID"
)

ggplot(pcoa_clr_df, aes(Axis.1, Axis.2, colour = DS)) +
  geom_point(size = 3) +
  scale_color_viridis_c() +
  theme_bw() +
  labs(
    title = "CLR-based PCoA (Aitchison Distance) — Dominance Rank",
    x = "PCoA1",
    y = "PCoA2"
  )


# PERMANOVA — set row names to match distance matrix labels
pcoa_clr_complete <- pcoa_clr_df %>%
  filter(!is.na(DS))
```

### PERMANOVA models
```{r}
# rank only
adonis2(
  dist_clr ~ DS + Group + age_years,
  data = pcoa_clr_complete %>% tibble::column_to_rownames("SampleID"),
  permutations = 9999
)

adonis2(
  dist_clr ~ DS + Group + age_years,
  data = pcoa_clr_complete %>% tibble::column_to_rownames("SampleID"),
  permutations = 9999,
  by = "margin"
)

# rank, group interaction
adonis2(
  dist_clr ~ DS * Group + age_years,
  data = pcoa_clr_complete %>% tibble::column_to_rownames("SampleID"),
  permutations = 9999,
)

adonis2(
  dist_clr ~ DS * Group + age_years,
  data = pcoa_clr_complete %>% tibble::column_to_rownames("SampleID"),
  permutations = 9999,
  by = "margin"
)
```
### Checking of assumptions
```{r}
# checking correlations
cor(
  pcoa_clr_complete %>%
    dplyr::select(
      DS,
      age_years
    )
)

# Checking homogenity of multivariate dispersion
disp_group <- betadisper(dist_clr,
                         pcoa_clr_complete$Group)

anova(disp_group)

permutest(disp_group)
```

## UniFrac Analysis Preparation
```{r}
# combine phyloseq to phylo tree
ps_tree <- merge_phyloseq(ps_pruned, phy_tree(tree$data))

ps_tree

```

## Weighted UniFrac
```{r}
dist_w <- phyloseq::distance(ps_tree, method = "wunifrac")
ordu_w <- ordinate(ps_tree, method = "PCoA", distance = dist_w)

plot_ordination(ps_tree, ordu_w) +
  geom_point(size = 4) +
  scale_color_viridis_c() +
  theme_bw() +
  labs(title = "Weighted UniFrac PCoA")

```

### PERMANOVA models
```{r}
adonis2(
  dist_w ~ DS + Group + age_years,
  data = as(sample_data(ps_tree), "data.frame"),
  permutations = 9999
)

adonis2(
  dist_w ~ DS + Group + age_years,
  data = as(sample_data(ps_tree), "data.frame"),  permutations = 9999,
  by = "margin"
)


adonis2(
  dist_w ~ DS * Group + age_years + DS + Group,
  data = as(sample_data(ps_tree), "data.frame"),  permutations = 9999,
)

adonis2(
  dist_w ~ DS * Group + age_years,
  data = as(sample_data(ps_tree), "data.frame"),
  permutations = 9999,
  by = "margin"
)

```

### Checking of assumptions
```{r}
# checking correlations
cor(
  as(sample_data(ps_tree), "data.frame") %>%
    dplyr::select(
      DS,
      age_years)
)

# Checking homogenity of multivariate dispersion
disp_group_b <- betadisper(dist_clr,
                         as(sample_data(ps_tree), "data.frame")$Group)

anova(disp_group_b)

permutest(disp_group_b)
```

# Alpha diversity matrices
## Running the Analyses
```{r}
# Calculate alpha diversity metrics
alpha_ranked <- estimate_richness(
  ps_pruned,
  measures = c(
    "Observed",
    "Shannon",
    "Simpson",
    "InvSimpson"
  )
) %>%
  tibble::rownames_to_column("sample") %>%
  mutate(
    Evenness = Shannon / log(Observed)
  )
```

### Merging diversity metrices to metadata
```{r}
# Merging diversity metrices to metadata
shannon_ssb <- alpha_ranked %>% 
  mutate(sample_key = gsub("\\.", "-", sample)) %>% 
  left_join(
    metadata  %>% 
      tibble::rownames_to_column("sample_key")  %>% 
      dplyr::select(c("sample_key", "monkey_id", "DS", "age_years", "Group", "adjusted_SSB", "adjusted_SSB_giver", "adjusted_SSB_receiver")),
    by = "sample_key"
  )
```

### Shannon diversity
```{r}
# Linear model for Shannon
lm_shannon <- lm(Shannon ~ DS + Group + age_years, data = shannon_ssb)
summary(lm_shannon)


# Linear model Shannon (rank), group interaction
lm_shannon_int <- lm(Shannon ~ DS * Group + age_years, data = shannon_ssb)
summary(lm_shannon_int)
```

### Evenness Index
```{r}
# Linear model for Evenness
lm_evenness <- lm(Evenness ~ DS + Group + age_years, data = shannon_ssb)
summary(lm_evenness)


# Linear model Evenness (rank), with group interaction
lm_evenness_int <- lm(Evenness ~ DS * Group + age_years, data = shannon_ssb)
summary(lm_evenness_int)
```

## Checking assumptions for Linear Models
```{r}
# Load packages
library(performance)
library(car)
library(lmtest)

# Store all models in a list
models <- list(
  Shannon = lm_shannon,
  Shannon_int = lm_shannon_int,
  Evenness = lm_evenness,
  Evenness_int = lm_evenness_int
)

# Run diagnostics for every model
for (name in names(models)) {

  cat("\n=========================================\n")
  cat(name, "\n")
  cat("=========================================\n\n")

  model <- models[[name]]

  ## Standard diagnostic plots
  par(mfrow = c(2, 2))
  plot(model)
  par(mfrow = c(1, 1))

  ## Variance Inflation Factors
  cat("\nVariance Inflation Factors\n")
  print(vif(model))

  ## Overall model diagnostics
  cat("\nPerformance diagnostics\n")
  print(check_model(model))

}
```

