---
title: "Rank Write-up Raw Values"
output: html_notebook
---

This document generates the number of aggression bouts used for each year

```{r}
rm(list = ls())

library(tidyverse)
library(lubridate)
```

# loading csv
```{r}
ag <- read.csv("~/Savolainen Lab/SSB/Data/Rank/Ag_Su_AdLib.csv")
```

# Removing unknown IDs
Unknown IDs cannot be used for rank determination
```{r}
exclude_ids <- c(
  "JVUN", "FUN", "JVFUN", "JVMUN",
  "MUN", "SAMUN", "INFANT"
)

ag_filtered <- ag %>%
  filter(
    !ACTOR1 %in% exclude_ids,
    !RECIP1 %in% exclude_ids
  )

ag_filtered <- ag_filtered %>%
  mutate(DATE = ymd(DATE))
```

# 2026 number of agonistic interactions
116 males, 502 agonistic interactions in 2026
```{r}
twosix <- read.csv("~/Savolainen Lab/SSB/Data/Rank/combined_cleaned_2026.csv")

nrow(na.omit(twosix))
n_distinct(na.omit(c(twosix$Winner, twosix$Loser)))
```


# 2023, 2024 number of agonistic interactions
151 males, 2573 agonistic interactions in 2023
132 males, 826 agonistic interactions in 2024
```{r}
twothree <- ag_filtered %>% 
  filter(year(DATE) == 2023) 
nrow(na.omit(twothree))
n_distinct(na.omit(c(twothree$ACTOR1, twothree$RECIP1)))
  
twofour <- ag_filtered %>% 
  filter(year(DATE) == 2024)
nrow(na.omit(twofour))
n_distinct(na.omit(c(twofour$ACTOR1, twofour$RECIP1)))

```

# total dataset
Total of 7643 agonistic interactions
478 unique males
```{r}
nrow(na.omit(ag_filtered))
n_distinct(na.omit(c(ag_filtered$ACTOR1, ag_filtered$RECIP1)))
```


# pre-2023
5810 agonistic interactions
411 males
```{r}
bef_2023 <- ag_filtered %>% 
  filter(year(DATE) < 2024)
nrow(na.omit(bef_2023))
n_distinct(na.omit(c(bef_2023$ACTOR1, bef_2023$RECIP1)))

```

