---
title: "For Thesis: MaAsLin2 (Generating facet plots)"
output: html_notebook
---

In this notebook, we will generate facet plots for immune-associated genera and pathogenic genera. 

# Preparing world environment
```{r}
rm(list = ls())

# loading packages
library(tidyverse)
library(ggplot2)
library(patchwork)
library(phyloseq)
library(qiime2R)
library(ggeffects)
library(vegan)
library(cowplot)
library(emmeans)
```

## Reading files
```{r}
# Change paths if needed
table_qza    <- "~/Savolainen Lab/Microbiome/Qiime/Outputs for PCoA/feature_table_decontam.qza"
taxonomy_qza <- "~/Savolainen Lab/Microbiome/Qiime/Outputs for PCoA/taxonomy_silva.qza"
metadata_tsv <- "~/Savolainen Lab/Microbiome/Qiime/Outputs for PCoA/metadata.tsv"

# Feature table
table <- read_qza(table_qza)$data

# Taxonomy
taxonomy <- read_qza(taxonomy_qza)$data

# Metadata
metadata <- read.table(
  metadata_tsv,
  header       = TRUE,
  sep          = "\t",
  row.names    = 1,
  comment.char = "",
  check.names  = FALSE
) %>%
  tibble::rownames_to_column("SampleID")
```

# Building the Phyloseq Object
```{r}
# OTU table — remove BLANK and CTR samples, taxa are rows in QIIME2 feature tables
exclude_samples <- colnames(table)[grepl("BLANK|CTR", colnames(table), ignore.case = TRUE)]
otu_data <- table[, !colnames(table) %in% exclude_samples]
otu <- otu_table(as.matrix(otu_data), taxa_are_rows = TRUE)

# Split taxonomy into ranks (separator is ";" with no space)
tax_split <- str_split_fixed(taxonomy$Taxon, ";", 7)
colnames(tax_split) <- c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species")
rownames(tax_split) <- taxonomy$Feature.ID

tax  <- tax_table(as.matrix(tax_split))

metadata_ps <- metadata %>%
  tibble::column_to_rownames("SampleID")

meta <- sample_data(metadata_ps)
ps <- phyloseq(otu, tax, meta)

# Filter: keep taxa present in >= 4 samples AND with total abundance >= 20
ps <- filter_taxa(ps, function(x) sum(x >= 1) >= 4 & sum(x) >= 20, prune = TRUE)

ps
```

## Basic checks
```{r}
nsamples(ps)
ntaxa(ps)
sort(sample_sums(ps))
```

## Remove low-depth samples
```{r}
ps_pruned <- prune_samples(sample_sums(ps) > 10000, ps)
ps_pruned
```

# Analyses
## Relative Abundance Transformation
```{r}
ps.rel <- transform_sample_counts(ps_pruned, function(x) x / sum(x))
```

## Preparing metadata file
```{r}
# Behavioural rates
rates <- read.csv("~/Savolainen Lab/SSB/Data/updatedssbrates.csv") %>% 
  mutate(
    Animal_ID = dplyr::recode(Animal_ID,
                            "5E+01" = "5E1")) %>% 
  rename(monkey_id = Animal_ID)
  

# Dominance rank
rank <- read.csv(
  "~/Savolainen Lab/SSB/Data/Rank/rank_all_gut.csv"
) %>%
  rename(monkey_id = Animal_ID) %>%
  mutate(
    monkey_id = dplyr::recode(
      monkey_id,
      "5.00E+01" = "5E1"
    )
  ) %>%
  dplyr::select(-X, -Year)

age <- read.csv("~/Savolainen Lab/Age/age_gut.csv") %>% 
  dplyr::select(-"X") %>% 
  rename(monkey_id = Animal_ID) %>% 
  mutate(
    monkey_id = dplyr::recode(
      monkey_id,
      "5EE1" = "5E1"
    )
  )

# Join behavioural metadata
metadata <- metadata %>%
  left_join(rates, by = "monkey_id") %>%
  left_join(rank, by = "monkey_id")

metadata <- metadata %>% 
  dplyr::select(-c("SSB_Giver", "SSB_Receiver", "DSB_Raw", "SSB_Raw", "X")) %>% 
  left_join(age, by = "monkey_id")

meta_new <- metadata %>% 
  drop_na(DS)
```

```{r}
ps.rel <- prune_samples(meta_new$SampleID, ps.rel)
ps_pruned <- prune_samples(meta_new$SampleID, ps_pruned)

# Keep only metadata for samples retained in the phyloseq object
meta_new <- meta_new %>%
  filter(SampleID %in% sample_names(ps_pruned)) %>%
  arrange(match(SampleID, sample_names(ps_pruned))) %>%
  tibble::column_to_rownames("SampleID")

# Replace sample metadata
sample_data(ps.rel) <- sample_data(meta_new)
sample_data(ps_pruned) <- sample_data(meta_new)

# Check that everything matches
nsamples(ps.rel)
nsamples(ps_pruned)
nrow(meta_new)
```

# Genus-level comparisons
## Agglomerate to genus and melt
```{r}
# Agglomerate counts to genus
ps.genus <- phyloseq::tax_glom(ps_pruned, taxrank = "Genus")

# Extract count matrix
otu <- as(otu_table(ps.genus), "matrix")

# Ensure samples are rows (required by decostand)
if (taxa_are_rows(ps.genus)) {
  otu <- t(otu)
}

# Add pseudocount to avoid zeros
otu <- otu + 1

# CLR transformation
otu.clr <- decostand(otu, method = "clr")

# Put CLR values back into phyloseq
otu_table(ps.genus) <- otu_table(t(otu.clr), taxa_are_rows = TRUE)

# Convert to data frame
genus.df <- psmelt(ps.genus)
```

## Editing metadata
```{r}
meta <- data.frame(sample_data(ps.genus))

meta$Group <- factor(meta$Group)

# Continuous predictors
meta$DS <- as.numeric(meta$DS)
meta$age_years <- as.numeric(meta$age_years)


```

## Making genus names visible in analyses
```{r}
# Extract genus names
genus_names <- as.character(tax_table(ps.genus)[, "Genus"])

# Replace missing names
genus_names[is.na(genus_names)] <- "Unclassified"

# Make duplicate names unique
genus_names <- make.unique(genus_names)

# Rename columns of otu
colnames(otu.clr) <- genus_names
```

# Preparing graphs
## Formatting data into suitable table
```{r}
plot_data <-
meta %>%
    bind_cols(as.data.frame(otu.clr))

plot_data <- plot_data %>%
  rename(
    Escherichia_Shigella = `Escherichia-Shigella`
  )
```


## Refitting models
### Refitting linear models immune-associated
```{r}
lm_bact_immune <- lm(Bacteroides ~DS * Group + age_years, data = plot_data)

lm_lach_immune <- lm(Lachnoclostridium ~ DS + Group + age_years, data = plot_data)
```

### Refitting linear models pathogenic
```{r}
lm_eschshi_path <- lm(Escherichia_Shigella ~ DS + Group + age_years, data = plot_data)

lm_strep_path <- lm(Streptococcus ~ DS + Group + age_years, data = plot_data)

```

## Using ggpredict to create predictions
```{r}

pred_bact <- ggpredict(
  lm_bact_immune,
  terms = c("DS", "Group")
)

pred_lach <- ggpredict(
  lm_lach_immune,
  terms = "age_years"
)

pred_esch <- ggpredict(
  lm_eschshi_path,
  terms = "DS"
)
```

# Plotting graphs
```{r}
theme_thesis <- theme_classic(base_size = 16) +
theme(
    axis.title = element_text(
        face = "plain",
        size = 22
    ),
    axis.text = element_text(
        size = 20
    ),
    axis.ticks.length = unit(0.22, "cm"),
    legend.key.size = unit(1.2, "lines"),
    legend.title = element_text(
        size = 20,
        face = "bold"
    ),
    legend.text = element_text(
        size = 18
    ),
    legend.position = "top",
    plot.margin = margin(
        8.0,
        8.0,
        8.0,
        8.0)
)
```

## Bacteroides interaction plot (immune)
```{r}
plot_bact <- ggplot() +

geom_point(
  data = plot_data,
  aes(
    DS,
    Bacteroides,
    colour = Group
  ),
  size = 3.2,
  alpha = 0.75
) +

  geom_ribbon(
  data = pred_bact,
  aes(
    x = x,
    ymin = conf.low,
    ymax = conf.high,
    fill = group,
    group = group
  ),
  alpha = 0.12,
  colour = NA,
  inherit.aes = FALSE,
  show.legend = FALSE
) +
  
geom_line(
  data = pred_bact,
  aes(
    x = x,
    y = predicted,
    colour = group),
  linewidth = 2
) +

scale_colour_manual(values = c(
  R = "#D55E00",
  V = "#0072B2"
)) +

labs(
  x = "David's Score",
  y = expression("CLR abundance (" * italic("Bacteroides") * ")"),
  colour = "Group"
) +

theme_thesis
plot_bact
```


## Escherichia-Shigella (pathogenic)
```{r}
plot_esch <-ggplot() +

geom_point(
  data = plot_data,
  aes(
    DS,
    `Escherichia_Shigella`
  ),
  size = 3.2,
  alpha = 0.75
) +

  geom_ribbon(
  data = pred_esch,
  aes(
    x = x,
    ymin = conf.low,
    ymax = conf.high
  ),
  alpha = 0.12,
  colour = NA,
  inherit.aes = FALSE
) +

geom_line(
  data = pred_esch,
  aes(x,
      predicted),
  linewidth = 2
) +
  
labs(
  x = NULL,
  y = expression(
    atop(
      "CLR abundance",
      paste("(", italic("Escherichia-Shigella"), ")")
      )
  )
) +

theme_thesis
plot_esch
```

### Combined four-panel figure (A–B)
```{r}
# Legend
legend <- get_legend(
  plot_bact +
    theme(
      legend.position = "bottom",
      legend.justification = "center"
    )
)

# Remove legends from plots
plot_esch <- plot_esch +
  theme(
    legend.position = "none",
    axis.title.x = element_blank(),
    plot.tag = element_text(face = "bold", size = 20)
  ) +
  labs(tag = "A")
plot_bact <- plot_bact +
 theme(
    legend.position = "none",
    plot.tag = element_text(face = "bold", size = 20)
  ) +
  labs(tag = "B")

# Assemble vertically
figure_microbiome <-
  plot_esch /
  wrap_elements(legend) /
  plot_bact +
  plot_layout(
    heights = c(0.9, 0.12, 1.1)
  ) & 
  theme(
    plot.margin = margin(5, 8, 5, 8)
  )

figure_microbiome

ggsave(
  "~/Savolainen Lab/Microbiome/Figures/facet_combined_AB_new.tiff",
  figure_microbiome, 
  width = 10,
  height = 12.5,
  dpi = 600,
  compression = "lzw"
)
```

# Reporting SE for interaction term
```{r}
emtrends(
  lm_bact_immune,
  specs = "Group",
  var = "DS"
)
```
